---
title: "RB_HMC_2_Diffusion_map"
author: "Austin Reed"
date: "01/09/2020"
output: html_document
---

```{r, message=FALSE}
library(ggplot2)
library(plyr)
library(dplyr)
library(diffusionMap)

#Bioc Packages
library(scran)
library(scater)
library(destiny)


sce.LumHMC <- readRDS("../../data/RB_vs_HMC/Luminal_HMC/sce_HMCsubC1.rds")
#sce.LumHMC
```

Selecting out the luminal class cells
```{r}
#as the previous analysis didn't appear to show much significant heterogeneity between the RB and HMC cells within the Milk bigcluster I will not subsection them for this analysis
#I tested this on sce.toy2 and it seemed to work fine there so it should work fine here too

#sce is your large data set

sce.LumHMC <- sce[,colData(sce)$Classes == 'Luminal_HMC']
sce.LP <- sce[,colData(sce)$Classes == 'Luminal_LP']
sce.HR <- sce[,colData(sce)$Classes == 'Luminal_HR']

Barcodes.LumHMC <- sample(sce.LumHMC$Barcode,1000)
Barcodes.LP <- sample(sce.LP$Barcode,1000)
Barcodes.HR <- sample(sce.HR$Barcode,1000)
Barcodes <- c(Barcodes.LumHMC,Barcodes.LP,Barcodes.HR)

sce.sub <- sce[,Barcodes]
sce.sub
table(sce.sub$Classes)
```

selecting big Cluster
```{r}
sce.LumHMC <- sce.LumHMC[,colData(sce.LumHMC)$big.clusters == 'cluster1']
colData(sce.LumHMC)
```

Considering only HVG's (This was an attempt at the method you sent but many functions had become defunct and I couldnt quite get it to work so I'll do my own method in the next code block)
```{r}
# m = as.matrix(logcounts(sce.LumHMC))
# 
# #trendVar() is defunct I will use fitTrendVar() instead
# means <- rowMeans(m)
# vars <- rowVars(m)
# fit <- fitTrendVar(means,vars, parametric=TRUE)
# decVar <- modelGeneVar(m,fit)
# ord.decVar <- decVar[order(decVar$bio,decreasing=TRUE),]
# hvg <- ord.decVar[1:(nrow(ord.decVar)/20),]
```

Gene var
```{r}
m = as.matrix(logcounts(sce.LumHMC))

dec <- modelGeneVar(sce.LumHMC)
hvg <-getTopHVGs(dec, prop = 0.1)
head(hvg)
```

DM construction (OLD)
```{r}
# set.seed(1)
# dm <- calculateDiffusionMap(sce.LumHMC, ncomponents=2,subset_row=hvg, dimred='PCA')
```


Generate map coordinates (think its now working)
```{r}
set.seed(6)
#Takes a *very*(5-10ish mins I think) long time on my computer so try with PCA
dm <- DiffusionMap(data=t(m[hvg,]), n_pcs = 50)
```


Prepare to plot
```{r}
colData(sce.LumHMC)$DC1 <- dm$DC1
colData(sce.LumHMC)$DC2 <- dm$DC2


# head(DC1)
# head(DC2)
```

Plot coloured by state
```{r}
p <- ggplot(as.data.frame(colData(sce.LumHMC)), aes(x=DC1, y=DC2, color=State)) +
  geom_point()

ggsave(filename="../../data/RB_vs_HMC/Luminal_HMC/Diffusion_map_RB_HMC.pdf",p,width=8,height=8)

p
```

Plot coloured by Cluster
```{r}
sample.set <- c("HMC1", "HMC2", "HMC3","HMC4", "RB1", "RB2","RB3","RB4")
sample.colours <- c("salmon1", "peachpuff2", "burlywood", "#F8766D", "steelblue3", "slategray3", "paleturquoise3","#00BFC4")

p.sample <- ggplot(as.data.frame(colData(sce.LumHMC)), aes(x=DC1, y=DC2, color=Sample)) + 
  geom_point() + scale_color_manual(values=setNames(sample.colours, sample.set)) +
  theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))
p.sample

p.cluster <- ggplot(as.data.frame(colData(sce.LumHMC)), aes(x=DC1, y=DC2, color=GraphClusters)) + geom_point() + theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))
p.cluster
```

Plot coloured by Sample
```{r}

```

Psuedotime analysis
```{r}
rootcell <- which.min(dm$DC2)

dpt <- DPT(dm, tips = rootcell)
plot(dpt)
```

```{r}
pdf("../../data/RB_vs_HMC/Luminal_HMC/dpt_HMCLuminalRestingBasal.pdf", width=8, height=8)
plot(dpt)

dev.off()
```

Return the session information
```{r}
sessionInfo()
```

