---
title: "RB_HMC_2_Differential_expression"
author: "Austin Reed"
date: "28/08/2020"
output: html_document
---

```{r, message=FALSE}
library(ggplot2)
library(plyr)
library(dplyr)
library(scatterplot3d)
library(pheatmap)

#Bioc Packages
library(scran)
library(scater)
library(iSEE)
library(rgl)
library(edgeR)
library(EnhancedVolcano)
library(topGO)
library(org.Hs.eg.db)
library(Rgraphviz)

sce.LumHMC <- readRDS("../../data/RB_vs_HMC/Luminal_HMC/sce_HMCsubC1.rds")
```

UMAP
```{r}
plotUMAP(sce.LumHMC, colour_by = 'GraphClusters')
plotUMAP(sce.LumHMC, colour_by='State')
```

Have a look at the clusters and types of cells in each
```{r}
table(sce.LumHMC$State,sce.LumHMC$GraphClusters)
```
Again but for sample
```{r}
table(sce.LumHMC$Sample,sce.LumHMC$GraphClusters)
```


I will then compare the 193 RB cells to a random selection of 193 HM cells from this BigCluster.
Collect the samples for comparison
```{r}
set.seed(100)
sce.HM <- sce.LumHMC[,colData(sce.LumHMC)$State=='HM']
HMbarcodes <- sample(colData(sce.HM)$Barcode, 193)

sce.RB <- sce.LumHMC[,colData(sce.LumHMC)$State=='RB']
RBbarcodes <- colData(sce.RB)$Barcode

barcodes <- c(HMbarcodes, RBbarcodes)

sce.Lum.mini <- sce.LumHMC[,barcodes]
sce.Lum.mini

# set.seed = 100
# 
# keep1 = mapply(function(x) { if (x=='RB') return(TRUE) else return(FALSE) }, sce.LumHMC$State)
# table(keep1)
# 
# sce.sample <- sce.LumHMC[,!keep1]
# keep <- sample(sce.sample$Barcode, 193)
# a <- !(keep1+!keep1)
# for (i in keep) {
#   a <- a + (mapply(function(x) { if (x==i) return(TRUE) else return(FALSE) }, sce.LumHMC$Barcode))
# }
# 
# keep2 <- as.logical(a)
# 
# sce.LumHMC$keep1 <- keep1
# sce.LumHMC$keep2 <- keep2
```


Then create aggregate summed vectors to compare
```{r}
summed <- aggregateAcrossCells(sce.Lum.mini,
                               ids=DataFrame(label=sce.Lum.mini$State,
                                             sample=sce.Lum.mini$Sample))
summed
table(summed$label,summed$sample)
```

Make DGEList object
```{r}
y <- DGEList(counts(summed), samples=colData(summed))
y
```
Pre-processing
```{r}
discarded <- isOutlier(y$samples$lib.size, log=TRUE, type='lower')
y <- y[,!discarded]
summary(discarded)
```

Removing lowly expressed genes
```{r}
keep <- filterByExpr(y,group=summed$State)
summary(keep)
y <- y[keep,]
```

Normalize
```{r}
y <- calcNormFactors(y)
y$samples
```
 

Stat modelling
```{r}
design <- model.matrix(~factor(State) , y$samples)
design
```

NB dispersion estimation
```{r}
y <- estimateDisp(y, design)
summary(y$trended.dispersion)
y
```

Plot biological coefficient of variation graph
```{r}
plotBCV(y)
```
Find QL neg. bin. fit
```{r}
fit <- glmQLFit(y, design, robust=TRUE)
summary(fit$var.prior)
summary(fit$df.prior)
```
 

Plot this
```{r}
plotQLDisp(fit)
```

Run test
```{r}
res <- glmQLFTest(fit, coef=ncol(design))
summary(decideTests(res))
```

See top gene markers
```{r}
topTags(res)
```

Adding false discovery rate (FDR) values to the res$table and saving this as a new object DEGs with the rows in order of the FDR
```{r}
res$table$FDR <- p.adjust(res$table$PValue, method="BH")
DEGs <- res$table[order(res$table$FDR),]
dim(DEGs) #1786x5
write.csv(DEGs, file= "../../data/RB_vs_HMC/Luminal_HMC/HMC_Cluster_DEGs.csv")
```
Volcano plot
```{r}
Volcano <- EnhancedVolcano(DEGs,
lab = rownames(DEGs),
x = 'logFC',
y = 'FDR',
title = 'DEGs RB in HMC cluster',
subtitle = '',
subtitleLabSize = 2,
legendPosition = "bottom",
pointSize = 3.0,
labSize = 2.0,
FCcutoff = 1,
pCutoff = 10e-2,
col = c("grey", "forestgreen", "steelblue", "red"),
#legendVisible = FALSE,
drawConnectors = FALSE,
typeConnectors = 'open')
Volcano

pdf("../../data/RB_vs_HMC/Luminal_HMC/HMC_Clusters_Volcano.pdf", width=12,height = 8)
Volcano
dev.off()
```

<!-- Summary table -->
<!-- ```{r} -->
<!-- sumry <- group_by(y$counts, y$State) %>% -->
<!--           summarise("Number of cells"=n(), -->
<!--           "UMIs"=median(sum), -->
<!--           "Genes Detected"=median(detected), -->
<!--           "HMC1"=count(Sample=="HMC1"), -->
<!--           "HMC2"=count(Sample=="HMC2"), -->
<!--           "HMC3"=count(Sample=="HMC3"), -->
<!--           "HMC4"=count(Sample=="HMC4")) -->
<!-- kable(sumry,caption="Summary of cluster content") -->
<!-- ``` -->

```{r}
coloured <- c("#00BFC4", "#F8766D")
Lum.clus <- c("RB", "HM")

sample.set <- c("HMC1", "HMC2", "HMC3","HMC4", "RB1", "RB2","RB3","RB4")
sample.colours <- c("salmon1", "peachpuff2", "burlywood", "#F8766D", "steelblue3", "slategray3", "paleturquoise3","#00BFC4")
 

#Genes with highest logFC c("MGP", "SFRP1", "CD74")
#Genes with lowest logFC c("DDIT4", "ATF3", "IER2")
#Genes associated with milk production c("LALBA", "CSN2", "CSN3")
Genes <- c("LALBA", "CSN2", "CSN3", "MGP", "SFRP1", "CD74", "DDIT4", "ATF3", "IER2")
myGraphs <- list()
pdf("../../data/RB_vs_HMC/Luminal_HMC/DEG_box_plots", width=12, height=8)
for (Gene in Genes) {

  m = logcounts(sce.Lum.mini)
  myGene <- as.numeric(m[Gene,])

 

  GeneOfInterest <- ggplot(as.data.frame(colData(sce.Lum.mini)), aes(x=State,y=myGene, fill=State)) +
    geom_jitter(aes(color=Sample), alpha=0.5, pch=19) +
    geom_boxplot(draw_quantiles=0.5) +
    ylab(Gene) +
  #  scale_y_log10() + 
    ggtitle(Gene) +
    theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) +
    scale_color_manual(values=setNames(sample.colours, sample.set)) + 
    scale_fill_manual(values=setNames(coloured,Lum.clus))
myGraphs[[Gene]] <- GeneOfInterest
print(myGraphs[[Gene]])
  }

 

dev.off()
```


```{r}
coloured <- c("#00BFC4", "#F8766D")
Lum.clus <- c("RB", "HM")

sample.set <- c("HMC1", "HMC2", "HMC3","HMC4", "RB1", "RB2","RB3","RB4")
sample.colours <- c("salmon1", "peachpuff2", "burlywood", "#F8766D", "steelblue3", "slategray3", "paleturquoise3","#00BFC4")
 

#Genes with highest logFC c("MGP", "SFRP1", "CD74")
#Genes with lowest logFC c("DDIT4", "ATF3", "IER2")
#Genes associated with milk production c("LALBA", "CSN2", "CSN3")
Genes <- c("LALBA", "CSN2", "CSN3", "MGP", "SFRP1", "CD74", "DDIT4", "ATF3", "IER2")
myGraphs <- list()
pdf("../../data/RB_vs_HMC/Luminal_HMC/DEG_box_plots2", width=12, height=8)
for (Gene in Genes) {

  m = counts(summed)
  myGene <- as.numeric(m[Gene,])

 

  GeneOfInterest <- ggplot(as.data.frame(colData(summed)), aes(x=State,y=myGene, fill=State)) +
    geom_jitter(aes(color=Sample), alpha=0.5, pch=19) +
    geom_boxplot(draw_quantiles=0.5) +
    ylab(Gene) +
    ggtitle(Gene) +
    theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) +
    scale_color_manual(values=setNames(sample.colours, sample.set)) + 
    scale_fill_manual(values=setNames(coloured,Lum.clus))
myGraphs[[Gene]] <- GeneOfInterest
print(myGraphs[[Gene]])
  }

 

dev.off()
```

----- Now Look at box plots for all Luminal HMC cells -----

```{r}
coloured <- c("#00BFC4", "#F8766D")
Lum.clus <- c("RB", "HM")

sample.set <- c("HMC1", "HMC2", "HMC3","HMC4", "RB1", "RB2","RB3","RB4")
sample.colours <- c("salmon1", "peachpuff2", "burlywood", "#F8766D", "steelblue3", "slategray3", "paleturquoise3","#00BFC4")
 

#Genes with highest logFC c("MGP", "SFRP1", "CD74")
#Genes with lowest logFC c("DDIT4", "ATF3", "IER2")
#Genes associated with milk production c("LALBA", "CSN2", "CSN3")
Genes <- c("LALBA", "CSN2", "CSN3", "MGP")
myGraphs <- list()
pdf("../../data/RB_vs_HMC/Luminal_HMC/DEG_box_plots3", width=12, height=8)
for (Gene in Genes) {

  m = logcounts(sce.LumHMC)
  myGene <- as.numeric(m[Gene,])

 

  GeneOfInterest <- ggplot(as.data.frame(colData(sce.LumHMC)), aes(x=State,y=myGene, fill=State)) +
    geom_jitter(aes(color=Sample), alpha=0.5, pch=19) +
    geom_boxplot(draw_quantiles=0.5) +
    ylab(Gene) +
    ggtitle(Gene) +
    theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) +
    scale_color_manual(values=setNames(sample.colours, sample.set)) + 
    scale_fill_manual(values=setNames(coloured,Lum.clus))
myGraphs[[Gene]] <- GeneOfInterest
print(myGraphs[[Gene]])
  }
 

dev.off()
```


PLotting all together (no separation by state)
```{r}
coloured <- c("#00BFC4", "#F8766D")
Lum.clus <- c("RB", "HM")

sample.set <- c("HMC1", "HMC2", "HMC3","HMC4", "RB1", "RB2","RB3","RB4")
sample.colours <- c("salmon1", "peachpuff2", "burlywood", "#F8766D", "steelblue3", "slategray3", "paleturquoise3","#00BFC4")
 

#Genes with highest logFC c("MGP", "SFRP1", "CD74")
#Genes with lowest logFC c("DDIT4", "ATF3", "IER2")
#Genes associated with milk production c("LALBA", "CSN2", "CSN3")
Genes <- c("LALBA", "CSN2", "CSN3", "MGP")
myGraphs <- list()
pdf("../../data/RB_vs_HMC/Luminal_HMC/DEG_box_plots4", width=10, height=7)
for (Gene in Genes) {

  m = logcounts(sce.LumHMC)
  myGene <- as.numeric(m[Gene,])

 

  GeneOfInterest <- ggplot(as.data.frame(colData(sce.LumHMC)), aes(x=Classes, y=myGene, sixe=1)) +
    geom_jitter(aes(color=Sample, size = State), alpha=0.5, pch=19) +
    ylab(Gene) +
    ggtitle(Gene) +
    theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) +
    scale_color_manual(values=setNames(sample.colours, sample.set)) 
myGraphs[[Gene]] <- GeneOfInterest
print(myGraphs[[Gene]])
  }
 

dev.off()
```

Segregate by Clusters
```{r}
coloured <- c("#00BFC4", "#F8766D")
Lum.clus <- c("RB", "HM")

sample.set <- c("HMC1", "HMC2", "HMC3","HMC4", "RB1", "RB2","RB3","RB4")
sample.colours <- c("salmon1", "peachpuff2", "burlywood", "#F8766D", "steelblue3", "slategray3", "paleturquoise3","#00BFC4")
 

#Genes with highest logFC c("MGP", "SFRP1", "CD74")
#Genes with lowest logFC c("DDIT4", "ATF3", "IER2")
#Genes associated with milk production c("LALBA", "CSN2", "CSN3")
Genes <- c("LALBA", "CSN2", "CSN3", "MGP")
myGraphs <- list()
pdf("../../data/RB_vs_HMC/Luminal_HMC/DEG_box_plots5", width=12, height=8)
for (Gene in Genes) {

  m = logcounts(sce.LumHMC)
  myGene <- as.numeric(m[Gene,])

 

  GeneOfInterest <- ggplot(as.data.frame(colData(sce.LumHMC)), aes(x=GraphClusters, y=myGene, sixe=1)) +
    geom_jitter(aes(color=Sample, size = State), alpha=0.5, pch=19) +
    ylab(Gene) +
    ggtitle(Gene) +
    theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) +
    scale_color_manual(values=setNames(sample.colours, sample.set)) 
myGraphs[[Gene]] <- GeneOfInterest
print(myGraphs[[Gene]])
  }
 

dev.off()
```

Scatter Graph showing the correlation between LALBA and CSN2 in Luminal HMC cells
```{r}
m=logcounts(sce.LumHMC)

sample.set <- c("HMC1", "HMC2", "HMC3","HMC4", "RB1", "RB2","RB3","RB4")
sample.colours <- c("salmon1", "peachpuff2", "burlywood", "#F8766D", "steelblue3", "slategray3", "paleturquoise3","#00BFC4") 

ggplot(as.data.frame(colData(sce.LumHMC)), aes(x=m['CSN2',], y=m['LALBA',], sixe=1)) +
    geom_jitter(aes(color=Sample), alpha=0.5, pch=19) +
    ylab('LALBA logcounts per cell') + xlab('CSN2 logcounts per cell') +
    ggtitle('LALBA CSN2 Correlation') +
    theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) +
    scale_color_manual(values=setNames(sample.colours, sample.set)) +
    geom_smooth(method = "lm", colour='black', formula=y~x) 
```

RB vs HMC gene counts
```{r}
coloured <- c("#00BFC4", "#F8766D")
Lum.clus <- c("RB", "HM")

ggplot(as.data.frame(colData(sce.LumHMC)), aes(x=State, y=sum)) +
    geom_jitter(aes(color=Sample), alpha=0.5, pch=19) +
    geom_violin(aes(fill=State)) +
    ylab('Total UMI counts per cell') +
    scale_y_log10() +
    ggtitle('UMI counts vs State') +
    theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) +
    scale_color_manual(values=setNames(sample.colours, sample.set)) +
    scale_fill_manual(values=setNames(coloured, Lum.clus))
```


RB vs HMC gene counts
```{r}
ggplot(as.data.frame(colData(sce.LumHMC)), aes(x=State, y=detected)) +
    geom_jitter(aes(color=Sample), alpha=0.5, pch=19) +
    geom_violin(aes(fill=State)) +
    ylab('Unique gene counts per cell') +
    scale_y_log10() +
    ggtitle('Unique gene expression vs State') +
    theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) +
    scale_color_manual(values=setNames(sample.colours, sample.set)) +
    scale_fill_manual(values=setNames(coloured, Lum.clus))
```


**GO term Analysis**

Determine Significant genes
```{r}
#All sig genes (Large FDR required as cells are very similar)
sig.genes <- subset(DEGs, DEGs$FDR < 0.15)
dim(sig.genes) #107

#Up-reg genes
up.genes <- subset(sig.genes, sig.genes$logFC > 0)
dim(up.genes) #29

#Down-reg genes
down.genes <- subset(sig.genes, sig.genes$logFC < 0)
dim(down.genes) #78
```

*GO term analysis of upregulated genes with biological process*

GO term annotations for RB cells using FDR corrected p.values and positive FC

```{r}
#generating named vector "geneList" which
geneList <- as.integer(rownames(res$table) %in% rownames(up.genes))
names(geneList) <- rownames(res$table)
count(geneList) #29
```

Setting all genes >0.01 (i.e. =1, part of our subset) as "significant"
```{r}
topDiffGenes <- function(allScore) {return(allScore > 0.01)}
x <- topDiffGenes(geneList)

sum(x) ## the number of selected genes -> 29
```

Creating the "topGOdata" object with at least 10 genes associated with each GO term
```{r}
GOdata <- new("topGOdata",
  description = "GO analysis of HMC cluster;RB vs HM",
  ontology = "BP",
  allGenes = geneList,
  geneSel = topDiffGenes,
  annot = annFUN.org,
  mapping = "org.Hs.eg",
  ID = "symbol",
  nodeSize = 10)
GOdata
```

To get the significant genes
```{r}
sg <- sigGenes(GOdata)
str(sg)
numSigGenes(GOdata)
#28 significant genes
```
Perform enrichment test, using the algorithum="weight01" (rather than algorithm = "classic") we take the GO hierarchy in to consideration
```{r}
resultFisher <- runTest(GOdata, algorithm = "weight01", statistic = "fisher")
allRes <- GenTable(GOdata, classicFisher = resultFisher, orderBy = "resultFisher", ranksOf = "classicFisher", topNodes = 50, numChar=200)
allRes
```
 
#add significant genes to the table
```{r}
allRes$genes <- sapply(allRes$GO.ID, function(x)
    {
      genes<-genesInTerm(GOdata, x)
      genes[[1]][genes[[1]] %in% sg] # myGenes is the queried gene list
    })
allRes$genes[which(allRes$topGO<0.0001)] # print those only with p-value < 0.05
allRes$genes <-vapply(allRes$genes, paste, collapse = ", ", character(1L))
#allRes<-subset(allRes, as.numeric(allRes[,"topGO"])<0.0001 | grepl("e-", allRes[,"topGO"]))

write.csv(allRes, file= "../../data/RB_vs_HMC/Luminal_HMC/RB_Upregulated_topGO.csv")
```

visualising the top nodes
```{r}
#pdf saved in scr HMC_vs_RB
showSigOfNodes(GOdata, score(resultFisher), firstSigNodes = 5, useInfo ='all')
GOdata_graph <- printGraph(GOdata, resultFisher, firstSigNodes = 5, fn.prefix = "tGO", useInfo = "all", pdfSW = TRUE)
GOdata_graph

#Finding genes of significant annotations
length(usedGO(GOdata))
#2927
```

Adding GeneRatio to the table which is number of significant genes annotated to the particular term divided by all terms used in the GO term analysis + making sure classidFisher registers as numeric for colour scales and ensuring "Term" is recognised not as a character so the plots will display the variables in order
```{r}
allRes$GeneRatio <- allRes$Significant/length(sigGenes(GOdata))
allRes$classicFisher <- as.numeric(allRes$classicFisher)
allRes$Term <- factor(allRes$Term, levels=allRes$Term)
allRes
```
 
Plotting GO terms as a bubble plot, where x-axis is GeneRatio, colour is classicfisher p-value and size is the number of genes significantly annotated to the term
```{r}
p <- ggplot(allRes,
      aes(x = GeneRatio, y = Term)) +
      geom_point(aes(size = Significant, color = classicFisher)) +
      theme_bw(base_size = 14)
p + scale_colour_gradientn(colours= c("#FEDF8F", "#D73027"))

ggsave(filename="../../data/RB_vs_HMC/Luminal_HMC/RB_Upregulated_Dotplot_GOterms_BP.pdf",width=11,height=7)
```

Plotting GO terms as a bubble plot, where x-axis is GeneRatio, colour is classicfisher p-value and size is the number of genes significantly annotated to the term Only top 10!
```{r}
p <- ggplot(allRes[1:10,],
      aes(x = GeneRatio, y = Term)) +
      geom_point(aes(size = Significant, color = classicFisher)) +
      theme_bw(base_size = 14)
p + scale_colour_gradientn(colours= c("#D73027","darkorange1"))

ggsave(filename="../../data/RB_vs_HMC/Luminal_HMC/RB_Upregulated_Dotplot_GOterms_top10_BP.pdf",width=11,height=4)
```
Plotting GO terms as a bargraph, where x-axis is GeneRatio, colour is classicfisher p-value
```{r}
q <- ggplot(allRes,
      aes(Term)) +
      ylab("GeneRatio") +
      geom_bar(aes(weight = GeneRatio, fill=classicFisher)) +
      coord_flip() +
      theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))

q + scale_fill_gradientn(colours= c("#FEDF8F", "#D73027"))

ggsave(filename="../../data/RB_vs_HMC/Luminal_HMC/RB_Upregulated_Bargraph_GOterms_BP.pdf",width=9,height=7)
```





```{r}
sessionInfo()
```


