---
title: "Testing_2"
author: "Austin Reed"
date: "26/08/2020"
output: html_document
---

Initial set up
```{r, message=FALSE}
library(ggplot2)
library(plyr)
library(dplyr)
library(scatterplot3d)
library(pheatmap)

#Bioc Packages
library(scran)
library(scater)
library(iSEE)
library(rgl)
library(edgeR)
library(EnhancedVolcano)

sce.toy2 <- readRDS("../data/sce_toy2.rds")

#also add in Alecia's cluster names found previously
Clusters <- c("C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8", "C9", "C10", "C11", "C12", "C13", "C14","C15")
Classes <- c("Luminal_HMC", "Luminal_HMC", "Luminal_HMC", "Luminal_HMC", "Luminal_HMC", "Immune", "Luminal_HMC", "Endothelial",  "Vascular_accessory", "Luminal_LP","Basal", "Luminal_HR", "Immune",  "Immune", "Fibroblast")
colData(sce.toy2)$Classes <- mapvalues(colData(sce.toy2)$GraphClusters, Clusters, Classes)

row.names(sce.toy2) <- rowData(sce.toy2)$Symbol
colData <- colData(sce.toy2)
rowData <- rowData(sce.toy2)
```

Colours for plotting
```{r}
#cluster colours
cluster.set <- c("C1", "C2", "C3","C4", "C5", "C6","C7","C8","C9", "C10", "C11","C12", "C13", "C14", "C15")
cluster.colors <- c("orchid", "plum", "#A58AFF", "slateblue2", "violet", "olivedrab", "#DF70F8", "lightgoldenrod", "sandybrown", "violetred3", "tomato1", "violetred2", "#99A800", "yellow4", "tan3")

#sample colours
sample.set <- c("HMC1", "HMC2", "HMC3","HMC4", "RB1", "RB2","RB3","RB4")
sample.colours <- c("salmon1", "peachpuff2", "burlywood", "#F8766D", "steelblue3", "slategray3", "paleturquoise3","#00BFC4")
```

Simple PCA plot
```{r}
#plotPCA(sce.toy2, colour_by = 'GraphClusters')

plot_PCA <- ggcells(sce.toy2, mapping=aes(x=PCA.1, y=PCA.2,color=GraphClusters)) + geom_point() + scale_color_manual(values = setNames(cluster.colors,cluster.set)) + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))
plot_PCA
```

Simple UMAP plot
```{r}
#colour by sample
plot_UMAP <- ggcells(sce.toy2, mapping=aes(x=UMAP.1, y=UMAP.2,color=Sample)) + geom_point() + scale_color_manual(values = setNames(sample.colours,sample.set)) + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))
plot_UMAP

#colour by cluster
plot_UMAP <- ggcells(sce.toy2, mapping=aes(x=UMAP.1, y=UMAP.2,color=GraphClusters)) + geom_point() + scale_color_manual(values = setNames(cluster.colors,cluster.set)) + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))
plot_UMAP
```

Moveable 3D scatter plot
```{r}
options(rgl.printRglwidget = TRUE)

scatter_colors <- as.factor(colData(sce.toy2)$GraphClusters)
levels(scatter_colors) <- cluster.colors
scatter_colors <- as.character(scatter_colors)

PCA_data_frame <- as.data.frame(reducedDim(sce.toy2, 'PCA'))

plot3d(PCA_data_frame$PC1, PCA_data_frame$PC2, PCA_data_frame$PC3, 
       xlab="PC1", ylab="PC2", zlab="PC3", col=scatter_colors)
```

Creating the sample and immune cluster average vectors
```{r}
sce.toy2.immune = sce.toy2[,colData(sce.toy2)$Classes == 'Immune']
sce.toy2.immune = sce.toy2.immune[,colData(sce.toy2.immune)$State=='RB']

#this will be a new sce class with previous data except now they will be clumped together for 
#more rigid statistical analysis for DE genes
#summed <- aggregateAcrossCells(sce.toy2.immune, ids=data.frame(sample = sce.toy2.immune$Sample, label = sce.toy2.immune$GraphClusters))
summed <- aggregateAcrossCells(sce.toy2.immune,
                               ids=DataFrame(label=sce.toy2.immune$GraphClusters,
                                             sample=sce.toy2.immune$Sample))
summed
table(summed$sample,summed$label)
```
Make DGEList object
```{r}
y <- DGEList(counts(summed), samples=colData(summed))
y
```
Pre-processing
```{r}
discarded <- isOutlier(y$samples$lib.size, log=TRUE, type='lower')
y <- y[,!discarded]
summary(discarded)
```

Removing lowly expressed genes
```{r}
keep <- filterByExpr(y,group=summed$GraphClusters)
summary(keep)
y <- y[keep,]
```

Normalize
```{r}
y <- calcNormFactors(y)
y$samples
```
 

Stat modelling
```{r}
design <- model.matrix(~factor(GraphClusters) , y$samples)
design
```

NB dispersion estimation
```{r}
y <- estimateDisp(y, design = design)
y
```

Plot biological coefficient of variation graph
```{r}
plotBCV(y)
```
Find QL neg. bin. fit
```{r}
fit <- glmQLFit(y, design, robust=TRUE)
summary(fit$var.prior)
summary(fit$df.prior)
```
 

Plot this
```{r}
plotQLDisp(fit)
```

Run test
```{r}
res <- glmQLFTest(fit, coef=ncol(design))
summary(decideTests(res))
```

See top gene markers
```{r}
topTags(res)
```

Adding false discovery rate (FDR) values to the res$table and saving this as a new object DEGs with the rows in order of the FDR
```{r}
res$table$FDR <- p.adjust(res$table$PValue, method="BH")
DEGs <- res$table[order(res$table$FDR),]
dim(DEGs) #441x5
write.csv(DEGs, file= "../data/RB_vs_HMC/Immune_cells/RB_Immune_comparisons_AllDEGs.csv")
```
Volcano plot
```{r}
Volcano <- EnhancedVolcano(DEGs,
lab = rownames(DEGs),
x = 'logFC',
y = 'FDR',
title = 'DEGs RB Comparison of Immune cells',
subtitle = '',
subtitleLabSize = 2,
legendPosition = "bottom",
pointSize = 3.0,
labSize = 2.0,
FCcutoff = 2,
pCutoff = 10e-4,
col = c("grey", "forestgreen", "steelblue", "red"),
#legendVisible = FALSE,
drawConnectors = FALSE,
typeConnectors = 'open')
Volcano

pdf("../data/RB_vs_HMC/Immune_cells/DEGs_Volcano.pdf", width=12,height = 8)
Volcano
dev.off()
```



