---
title: "Testing"
author: "Alecia" #my supervisor began this and helped me through as my first trial run
date: "`r Sys.Date()`"
output: html_document
---

```{r, message=FALSE}
library(ggplot2)
library(plyr)
library(dplyr)
library(scatterplot3d)
library(pheatmap)
library(magick)

DataList <- readRDS("../data/toydata.rds")
counts <- DataList$counts
colData <- DataList$phenoData
rowData <- DataList$featureData
```
#
# BiocManager install
# ```{r, message=FALSE}
# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install(version = "3.11")
# ```
# 
# Install Bioc libraries
# ```{r, message=FALSE}
# BiocManager::install('iSEE')
# BiocManager::install('rgl')
# ```

Load BiocManager libraries
```{r, message=FALSE}
library(scran)
library(scater)
library(iSEE)
library(rgl)
```


```{r}
counts[1:3,1:5]
```
To look at the structure of a data frame, we can use the function "head"
```{r}
head(colData)
head(rowData)
```

Test plot
```{r}
LibrarySize <- ggplot(colData, aes(x=Sample,y=UmiSums, fill=Sample)) +
    geom_violin(draw_quantiles=0.5)+
    scale_y_log10() +
    ylab("Total UMI counts") +
    ggtitle("UMI counts across the Samples") +
    theme_bw()
LibrarySize
pdf("../data/lIBRARYsIZE.pdf", width=12, height=8)
LibrarySize
dev.off()
```

Forming the sce
```{r}
counts_matrix = as.matrix(counts)
sce.toy <- SingleCellExperiment(assays = list(logcounts = counts_matrix), rowData = rowData, colData = colData)
sce.toy
```


Gene Variation and HVG
```{r}
dec <- modelGeneVar(sce.toy)
hvg <-getTopHVGs(dec, prop = 0.1)
```

#set seed

Dimension reduction
```{r}
set.seed(10)
#I know this is a lot of components but its a small dataset so i just let it go
#I thought it might help with TSNE and UMAP but really it probably makes little difference.
sce.toy <- runPCA(sce.toy, ncomponents = 50, subset_row = hvg)
sce.toy <- runUMAP(sce.toy, dimred = 'PCA', n_neighbors = 5, min_dist = 0.9)
sce.toy <- runTSNE(sce.toy, dimred = 'PCA')
```

Clustering
```{r}
#note lower k gives more clusters.
set.seed(100)
NN_graph <- buildSNNGraph(sce.toy, use.dimred='PCA', k=6)
clusters <- igraph::cluster_louvain(NN_graph)$membership
colData(sce.toy)$label <- factor(paste0('c', clusters))
table(clusters)
```

Choosing colours for plots
```{r}
sample.set <- c('HMC', 'RB1', 'RB2', 'RB3')
sample.clusters <- as.character(levels(colData(sce.toy)$label))
sample.colors <- c('goldenrod1', 'darkslategray', 'darkseagreen1', 'dodgerblue')
cluster.colors <- c('green3','darkseagreen3','darkslategray1','darkorchid1','darkorange','dodgerblue1','goldenrod1','dodgerblue4','red3','cyan2','blue3', 'red','darkmagenta')
sample.clusters

```

Plotting first two PCA's
```{r}
#original plotting
plot_PCA_1 <- plotPCA(sce.toy, colour_by = 'label')

#ggplot with color by sample
plot_PCA_2 <- ggcells(sce.toy, mapping=aes(x=PCA.1, y=PCA.2,color=Sample)) + geom_point() + scale_color_manual(values = setNames(sample.colors,sample.set)) + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))

#ggplot with manual color by cluster (milk warm colors, tissue cold colors)
plot_PCA_3 <- ggcells(sce.toy, mapping=aes(x=PCA.1, y=PCA.2,color=label)) + geom_point() + scale_color_manual(values = setNames(cluster.colors,sample.clusters)) + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))

plot_PCA_2
plot_PCA_3
```

Plotting clusters on UMAP dimred
```{r}
#orginal plot
plot_UMAP <- plotUMAP(sce.toy, colour_by='label')

#colored by contour wrt sample batch
plot_UMAP_2 <- ggcells(sce.toy, mapping=aes(x=UMAP.1, y=UMAP.2,color=label)) + geom_point() + scale_color_manual(values = setNames(cluster.colors,sample.clusters)) + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))

plot_UMAP_2
```

PLotting clusters on TSNE dimred
```{r}
#orignal plot
plot_TSNE_1 <- plotTSNE(sce.toy, colour_by= 'label')

#colored by cluster wrt sample batch
plot_TSNE_2 <- ggcells(sce.toy, mapping=aes(x=TSNE.1, y=TSNE.2,color=label)) + geom_point() + scale_color_manual(values = setNames(cluster.colors,sample.clusters)) + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))

plot_TSNE_2
```

3D PCA plot
```{r}
#get color list to match scatterplot3d form
scatter_colors <- as.factor(colData(sce.toy)$label)
levels(scatter_colors) <- cluster.colors
scatter_colors <- as.character(scatter_colors)

#get points in correct data form
PCA_data_frame <- reducedDim(sce.toy,'PCA')
PCA_data_frame <- as.data.frame(PCA_matrix)

scatterplot3d(x=PCA_data_frame$PC1, y=PCA_data_frame$PC2, z=PCA_data_frame$PC3, color=scatter_colors, angle=60, xlab='PC1', ylab='PC2', zlab='PC3')
```

Moveable 3D scatter plot
```{r}
options(rgl.printRglwidget = TRUE)
 
plot3d(PCA_data_frame$PC1, PCA_data_frame$PC2, PCA_data_frame$PC3, 
       xlab="PC1", ylab="PC2", zlab="PC3", col=scatter_colors)
```

Save moving gif
```{r}
plot3d(PCA_data_frame$PC1, PCA_data_frame$PC2, PCA_data_frame$PC3, 
       xlab="PC1", ylab="PC2", zlab="PC3", col=scatter_colors)

play3d( spin3d( axis = c(0, 0, 1), rpm = 3), duration = 1)

# Save like gif
movie3d(
  movie="3dAnimatedScatterplot", 
  spin3d( axis = c(0, 0, 1), rpm = 3),
  dir='../data',
  duration = 30, 
  type = "gif", 
  clean = TRUE
)
```

Create Gene Markers list
```{r}
markers.toy <- findMarkers(sce.toy)
```

Generate set of top ten upregulated markers
```{r}
#not sure if this is the most efficient way to do this but it worked
#I wasnt able to get lapply() to work for some reason?

#I would also like to try to get the symbol as opposed to ID.?

top_10 = list(1:13)
for (i in 1:13) {
    inter <- markers.toy[[paste0('c',i)]]
    interesting_genes_c <- inter[inter$Top<=10 & inter$summary.logFC>0,]
    top_10[[i]] <- interesting_genes_c
}
```

Plot heatmap for top 10 marks for c1
```{r}
logFC_c1 <- getMarkerEffects(top_10[[1]])
pheatmap(logFC_c1, breaks=seq(-5, 5, length.out=101))
```
Replace id names with symbol names
```{r}
c1_marker_id <- top_10[[1]]
c1_marker_id <- rownames(c1_marker_id)
sce.rowData <- rowData(sce.toy)
values <- base::match(c1_marker_id, sce.rowData[,1])
c1_marker_symbol <- sce.rowData[values,2]

#fix NA one(s)
values <- base::match(NA, c1_marker_symbol)
c1_marker_symbol[values] <- c1_marker_id[values]

rownames(top_10[[1]]) <- c1_marker_symbol
```

Plot heatmap again with symbols this time
```{r}
logFC_c1 <- getMarkerEffects(top_10[[1]])
pheatmap(logFC_c1, breaks=seq(-5, 5, length.out=101))
```


